In-silico molecular enrichment and clearance of the human intracranial space

### 2024-12-19 paper/supplementary_main.tex
Make description of dispersion very structured.

\mer{MER: Stopped here.}

\subsection{Modelling dispersion in the CSF spaces}
\label{sec:app:dispersion}

\mer{MER: Moved the below from Results.}
\mar{To examine cardiac-driven dispersion, we compute a steady CSF flow field at peak systolic blood inflow. We mimic the resulting reduction of CSF space volume by imposing an inflow of $6\,$ml/s \cite{baledent2014imaging, causemann2022human} and $0.31\,$ml/s \cite{vinje2019respiratory} across the upper arachnoid mater ($\Gamma_{\rm AM-U}$) and the lateral ventricle ($\Gamma_{\rm LV}$) surface, respectively. We allow free outflow into the spinal SAS ($\Gamma_{\rm SSAS}$). In addition, we upscale the obtained pressure field to account for the lack of pulsatile inertial effects in our model.}

@@

CSF pulsates in the SAS and ventricular system with each cardiac and respiratory cycle, leading to an increase in effective solute diffusivity via a process known as Taylor dispersion \cite{taylor1953dispersion, watson1983diffusion}. Previous estimates of the magnitude of this effect in the CSF-filled space range from an enhancement factor of 5.8 \cite{keith2019dispersion} up to $10^4$ \cite{ray2021quantitative}. The large discrepancies in reported estimates can (at least partly) be attributed to different assumptions on representative values for local width, pressure differences and fluid velocities in the SAS, which reflect the diverse flow characteristics in the different regions of the intracranial CSF-filled space.

>>>>

\subsection{Estimating dispersion factors from pulsatile CSF flow}
\label{sec:app:dispersion}

The cardiac and respiratory cycles (typically 0.25--1.0 Hz in humans)
induce pulsatile flow of CSF in the ventricular system and in the
cranial and spinal SAS. Pulsatile flow leads to dispersion which in
turn may enhance molecular transport~\cite{taylor1953dispersion,
  watson1983diffusion, asgari2016glymphatic, keith2019dispersion,
  ray2021quantitative, troyetsky2021dispersion}. To account for the
dispersive effects over a longer time scale (hours to days) and with
spatial differences, we propose the following four-step approach for
the cardiac and respiratory contributions separately.
\begin{enumerate}%[\label=\roman]
\item
  To account for viscous forces, we compute spatially-varying CSF
  pressure fields $p^0_c, p^0_r$ in $\Omega_{\rm CSF}$ corresponding to the
  Stokes flow induced by the peak volumetric reduction of the CSF
  space in the respective cycle. This step is described in further
  detail below.
\item
  Next, we estimate the Womersley numbers $\alpha_c, \alpha_r$
  associated with the cardiac and respiratory flow patterns by the
  definition
  \begin{equation}
    \alpha^2_i = \frac{h^2 \omega_i \rho}{\mu} \qquad i \in \{c, r\},
  \end{equation}
  with CSF density $\rho = 10^3$ kg/m$^3$, a mean CSF space width $h =
  1.5$mm, and CSF viscosity $\mu$ given in~\Cref{tab:parameters}.  For
  the cardiac cycle, we consider an angular frequency $\omega_c = 2
  \pi$, while for the respiratory cycle, we set $\omega_r = 0.5
  \pi$. The resulting (square) Womersley numbers are $\alpha_c^2 = 20.2$
  for the cardiac cycle and $\alpha_r^2 = 5.05$ for the respiratory
  cycle.
\item
  To account for inertial forces in addition to the viscous forces, we
  use the respective Womersley number $\alpha_i$ to calculate an
  upscaled pressure field $p^1$ from $p^0$:
  \begin{equation}
    p^1_i(x) = \left ( 1 + \frac{\alpha_i^2}{8} \right ) p_i^0(x)
    \qquad \text{for } x \in \Omega_{\rm CSF} .
  \end{equation}
\item
  Further, assuming unsteady dispersion, we follow Sharp et
  al.~\cite{keith2019dispersion} to estimate local enhancement factors
  $S_i$ from the non-dimensionalized pressure gradients:
  \begin{equation}
    S_i = \frac{1}{\alpha_i^3} \frac{1}{\omega_i \mu/h} | \nabla p_i^1 |.
  \end{equation}
\item
  Finally, we account for the non-local nature of dispersion by
  smoothing ...
\end{enumerate}

\mar{To examine cardiac-driven dispersion, we compute a steady CSF flow field at peak systolic blood inflow. We mimic the resulting reduction of CSF space volume by imposing an inflow of $6\,$ml/s \cite{baledent2014imaging, causemann2022human} and $0.31\,$ml/s \cite{vinje2019respiratory} across the upper arachnoid mater ($\Gamma_{\rm AM-U}$) and the lateral ventricle ($\Gamma_{\rm LV}$) surface, respectively. We allow free outflow into the spinal SAS ($\Gamma_{\rm SSAS}$). In addition, we upscale the obtained pressure field to account for the lack of pulsatile inertial effects in our model.}

@@
